The weight of the mountains: Constraints on tectonic stress, friction, and fluid pressure in the 2008 Wenchuan earthquake from estimates of topographic loading

### 2015-02-13 revisions/rebuttal2.md

### Abstract

**CA.1**

> Do they mean the whole fault was verging on failure at the time of the
earthquake, or only part of it?

We have changed the relevant sentence to "We further estimate the static
friction and pore fluid pressure assuming that the fault was*, on average,* at
Mohr-Coulomb failure at the time of the Wenchuan earthquake."

@@

Additionally, when we did the log transformation of the y-axis of Figure 10,
we used the base-*e* log, not base-10; the anti-log of 3 in this case is ~20,
which is far more reasonable value than 1000. We have updated the text and
figure captions so that 'log' now reads 'ln' to avoid this confusion.

Nonetheless, we have reviewed our code and found a slight error in how
the optimal slip plane was calculated; essentially, the algorithm
assumed that dip slip faulting would be favored (where the optimal
plane strikes perpendicular to the trend of $\sigma_1$) but produces
an incorrect orientation otherwise.  We have fixed the bug, which has
changed these results slightly (the bug only affected these
calculations; that particular function was not used in the rest of the
study). The new Figure 10 shows that more of the data plot in the
favorability region for optimal faulting than previously; however,
most of the data (with low friction values) still plot in the
favorability region for the Wenchuan faults.

We have also added some discussion of how the calculated ratios relate to the
parameters:

"It is clear that both $T^\prime_{xx}$ and $\mu$ are strongly correlated with
$\mathrm{CFR}_o / \mathrm{CFR}_f$, and $\phi$ is to a lesser degree; nonetheless,
all significantly affect the relative ease of faulting on the Wenchuan faults
versus optimal faults. In particular, lower values for any of them favor slip
on the Wenchuan faults. The lowest $\mathrm{CFR}_o / \mathrm{CFR}_f$ value
($\sim$ 0.057, $\sim$ -2.9 in log space ) corresponds to the lowest value of
$\mu$ ($\sim$ 0.038), and is approximately the ratio of $\mu$ in the model to
0.6."

@@

Yes, $\sigma^B$ should be $M^B$, and $F_h,y$ instead of $F_h,yz$.

**CAp4**

> In equation (5) why are all three terms on the right side of the equal sign
in the first row multiplied by dh/dx? I can see why the ρgh term is, because if
the surface has a slope, then the slope will impart a horizontal shear stress.
I don’t understand why Mxx and Txx are multiplied by dh/dx though, because Mxx
and Txx already act horizontally. Similarly, why are Mxy and Txy multiplied by
dh/dy in the second row of the equation? Analogous comments apply to equation
(6). An explanation of this would be helpful.

Liu and Zoback [1992] devote some attention to the horizontal loading
function (primarily, see the "Effect of Topography on Tectonic Stress"
Section in their paper). While we do state in the beginning of the
section that we use methods derived by Liu and Zoback [1992], we
realized that we should also cite their paper where we give the
horizontal loading function, and we have edited the sentence equation
(5) to state that these loading functions are derived by Liu and
Zoback [1992].

Yes, Mxx, Txx, Mxy, and Txy act horizontally, but they represent
normal tractions in horizontal directions on vertical planes, and not
tractions on the free surface. Mxx, etc, above the reference level,
here taken to be mean sea level, only excite shear traction on the
free surface (warranting their inclusion into the horizontal loading
function) if topography varies. We think of this as due to the fact
that you are changing the total size of the vertical planes that
tractions resulting from Mxx, etc, are acting upon. This change in
horizontal normal tration results in a shear stress on the reference
surface and stresses are propagated into the half-space via the
Cerruti solutions.

### 2015-02-13 revisions/rebuttal2.md

This document contains the comments from the reviewer, as well as our changes to them.

@@

Yes, this is what we meant.  We have replaced 'fault' with 'earthquake' in this
sentence.

**C4.4a**

> (265) The phrase "posteriors of T" could have unanticipated (and undesirable)
reactions from some readers. I suggest using the alternative phrase "these
functions".

The use of the word 'posterior' in statistics is well-established and
is fairly commonly used in geophysical research papers that utilize
Bayesian estimation or inference. Additionally, as described in
section 4.2, our sampling estimation strategy results in samples of
some unknown posterior probability density function, this is the
solution to the inverse problem in a Bayesian sense, and so referring
to these collections of posterior samples as "functions" would be
misleading.

### 2015-03-18 revisions/styron_hetland_wenchuan_topo_revised.tex

%\begin{figure*}[ht!] \includegraphics[width=40pc]{../figures/lms_map.pdf}
%    \caption{Map of eastern Tibet and the Sichuan basin, showing active
%        structures from \citet{styron2010}. Faults that ruptured in the 2008
%        Wenchuan earthquake are shown in pink. GPS velocities are relative to
%        the mean velocity of sites within Sichuan basin, with 1$\sigma$
%        uncertainty, from the dataset of Liang et al.  \citet{liang2013}.
%        Beachball is from the Global CMT focal mechanism solution for the 2008
%    Wenchuan earthquake. BF = Beichuan fault. PF = Pengguan fault. P = Pengguan
%massif. Grey box shows the extent of Figures 3 and 9.} \label{fig:lms_map}
%\end{figure*}

@@

%\begin{figure*}[t!] \centering
%    \includegraphics[width=40pc]{../figures/topo_stress_possibilities.pdf}
%    \caption{Scenarios for topographic effects on rangefront thrust faulting.
%        (a) Topographic stresses promote thrust faulting. (b) Topographic
%        stresses inhibit thrust faulting.} \label{fig:topo_fault_scenarios}
%    \end{figure*}

@@

%\begin{table}
%\centering
%\begin{tabular}{l c l}
%\hline
%Parameter & Value & Unit \\
%\hline
%horizontal spacing & 851 & m \\ 
%vertical spacing & 1000 & m \\ 
%minimum depth & 851 & m (below sea level) \\ 
%maximum depth & 35851 & m (below sea level) \\ 
%density ($\rho$) & 2700 & kg m$^{-3}$ \\ 
%g & 9.81 & m s$^{-2}$ \\ 
%Green's function radius & 9e5 & m \\ 
%Poisson ratio & 0.25 & - \\ 
%\hline
%\end{tabular}
%\caption{Parameters for numerical calculations of topographic stresses.}
%\label{table:convo_params}
%\end{table}

@@

%\begin{figure}[t]
%\centering
%\includegraphics[width=20pc]{../figures/lms_topo_stresses_rot.pdf}
%\caption{Horizontal topographic stresses in the Longmen Shan region at 5
%km depth: black and red lines signify most and least compressive
%horizontal stresses, respectively. Other symbols are the same as in Figure
%1. Stresses shown are down-sampled from the
%discretization used in the calculations by a factor of six.}
%\label{fig:lms_topo_stresses_rot}
%\end{figure}

@@

%\begin{figure*}[htbp]
%\centering
%\includegraphics[width=40pc]{../figures/fault_stress_3d.pdf}
%\caption{Southwest-looking views of topographic stresses and coseismic
%slip on selected slip models. All views share same lateral extent, but
%the perspectives for B and C are more inclined than A. (A) $\sigma_n^M$
%(colors), slip magnitude (contours, 1 m interval) and hanging-wall
%topography on the \citet{feng2010} model of the Beichuan
%fault. Note the suppression of fault slip where normal stress is
%highest, such as below the Pengguan massif (P). Fault and topography
%share the same scale, with no vertical exaggeration. (B) $\tau_d^M$ (colors)
%and dip slip (contours, 1 m interval) \citet{qi2011}
%`rough' slip model of the Beichuan fault. (C) $\tau_s^M$ (colors) and strike
%slip (contours, 1
%m interval) on the \citet{qi2011} `rough' slip model if the Beichuan fault.}
%\label{fig:fault_stress_3d}
%\end{figure*}

@@

%\begin{figure}%[htbp]
%\centering
%\includegraphics[width=20pc]{../figures/feng_slip_sig_n_scatter.pdf}
%\caption{Coseismic slip magnitude and $\sigma^M_n$ on the four segments
%of the \citet{feng2010} coseismic slip model. Trendlines are
%L1 regressions and do not include points with no slip. ``NE B'' =
%Northeastern Beichuan fault. ``C B'' = central Beichuan fault. ``SW B'' =
%Southwestern Beichuan fault. ``P'' = Pengguan fault.}
%\label{fig:feng_slip_sig_n_scatter}
%\end{figure}

@@

%\begin{figure*}%[htbp]
%\centering
%\includegraphics[width=40pc]{../figures/T_scatters.pdf}
%\caption{Samples of
%$p(T^\prime_{\mathrm{max}},T^\prime_{\mathrm{min}} | D)$ associated with
%each of the coseismic slip models we consider, along with marginals of
%$T^\prime_{\mathrm{max}}$ and $T^\prime_{\mathrm{min}}$. Inset rose
%diagrams are histograms of azimuth of
%$T^\prime_{max}$.}
%\label{fig:T_scatters}
%\end{figure*}

@@

%\begin{figure*}[htbp]
%\centering
%\includegraphics[width=40pc]{../figures/mu_phi_fms.pdf}
%\caption{Samples of $p(\mu,\phi | D)$ for each coseismic slip model.
%Colors indicate magnitude of $T^\prime_{\mathrm{max}}$. Contour lines
%indicate relative density (i.e., likelihood) of posteriors (darker lines
%signify higher densities).}
%\label{fig:mu_phi_scatters}
%\end{figure*}

@@

%\begin{figure*}[t]
%\centering
%\includegraphics[width=40pc]{../figures/joint_pdfs.pdf}
%\caption{(a) Samples of
%$p_{\mathrm{J}}(T^\prime_{\mathrm{max}},T^\prime_{\mathrm{min}} | D)$,
%along with marginals of $T^\prime_{\mathrm{max}}$ and
%$T^\prime_{\mathrm{min}}$. (b) Samples of
%$p_{\mathrm{J}}(T^\prime_{\mathrm{\max}}, T^\prime_{\mathrm{\min}}/T^\prime_{\mathrm{max}} | D)$,
%along with marginals of $T^\prime_{\mathrm{max}}$ and
%$T^\prime_{\mathrm{min}} / T^\prime_{\mathrm{max}}$. (c) Histogram of
%azimuths of $T^\prime_{\mathrm{\max}}$. (d) Samples of
%$p_{\mathrm{J}}(\mu, \phi | D)$, with marginals, where color of the samples
%indicate magnitudes of $T^\prime_{\mathrm{\max}}$ and contour
%lines indicate relative density (i.e., likelihood) of posteriors (darker
%lines signify higher densities).}
%\label{fig:joint_posteriors}
%\end{figure*}

@@

%\begin{figure}[t]
%\centering
%\includegraphics[width=20pc]{../figures/lms_map_stresses_rot.pdf}
%\caption{Topographic and tectonic horizontal stresses (taken from the
%most likely estimates of $p(T|D)$ in the Wenchuan rupture region (black
%and red crosses) with horizontal maximum stress orientation data taken
%from before the 2008 Wenchuan event from the World Stress Map
%\citep{heidbach2009} (purple arrows), and horizontal maximum stress
%orientation data from after the earthquake at the WFSD-1 drill hole
%\citep{cui2014} (blue arrows). Other symbols are as in Figure
%\ref{fig:lms_map}. Stresses shown are downsampled from our computational 
%grid resolution by a factor of nine.}
%\label{fig:lms_stress_map}
%\end{figure}

The magnitudes of $T_{\mathrm{max}}$ and $T_{\mathrm{min}}$ are
dominantly constrained on the low end by our analysis, which is apparent
by the sharp decrease in the frequency of $p(T_{\mathrm{max}}|D)$ below
about $T^\prime_{\mathrm{max}} = 0.5$ (Figure~\ref{fig:joint_posteriors}). 
These results indicate that $T_{\mathrm{max}}$ of at
least $\sim$13.25 MPa km$^{-1}$ is necessary to overcome
topographic stresses resisting reverse and right-lateral slip on the
faults. We find that the highest likely ratio of strike-slip to dip-slip
shear along the Wenchuan earthquake faults is close to 1. This is
similar to the inferences of strain accumulation rates inferred from
`squishy' block modeling by \citet{loveless2011}, who
concluded that the rate of slip deficit accumulation in the thrust and
dextral senses were approximately equal. It should be noted that the
tectonic stresses we estimate here represent the accumulated stresses
prior to the Wenchuan earthquake, and the relation of these stresses to any
accumulated slip deficit needs to be through a model of strain
accumulation.

@@

%\begin{figure}%[tb]
%\centering
%\includegraphics{../figures/cfr_plots.pdf}
%\caption{Comparison of Coulomb failure ratio (CFR) on the Beichuan fault
%from the \citet{zhang2011} coseismic slip model to CFR on an
%optimally-oriented fault with $\mu = 0.6$, versus estimated $\mu$ on the
%Beichuan fault. Values less than 0 (1 in linear space) indicate that
%slip is favored on the Beichuan fault, even if it is not optimally
%oriented. Values are calculated for each point in the slip model for
%1000 randomly-drawn samples from $p(T,\mu,\phi |D)$.}
%\label{fig:cfr_ratios}
%\end{figure}
%
%\begin{table}[htb]
%\centering
%\begin{tabular}{l c c c c}
%\hline
%Parameter & Coef. & Std. Err. & t statistic & $P>|t|$ \\
%\hline
%Intercept & -0.2566 & 0.0020 & -128.8 & \textless{}0.0001 \\ 
%$\mu$ & 1.1168 & 0.0086 & 130.6 & \textless{}0.0001 \\ 
%$T^\prime_{xx}$ & 0.9138 & 0.0048 & 191.2 & \textless{}0.0001 \\ 
%$\phi$ & 0.5762 & 0.0055 & 105.7 & \textless{}0.0001 \\ 
%\hline
%\end{tabular}
%\caption{Sensitivity of CFR ratios to relevant stress state parameters:
%Results of multivariate linear regression of
%$\mathrm{CFR}_o/ \mathrm{CFR}_f)$ against $\mu$, $\phi$ and
%$T^\prime_{xx}$.}
%\label{table:cfr_regress}
%\end{table}

@@

\begin{acknowledgements}
All data used in this work are from published sources: \citet{shen2009};
\citet{feng2010}; \citet{qi2011}; \citet{zhang2011}; \citet{fielding2013}. SRTM
data are from CGIAR-SRTM: srtm.csi.cgiar.org. All code used is available at
\url{https://github.com/cossatot/halfspace} and
\url{https://github.com/wenchuan_topo_stress}. We thank the editor, Paul
Tregoning, and Steve Marshak and an anonymous reviewer for reviews. Funding for
this study was provided by the University of Michigan. Joe Kington wrote the
code to plot Figure \ref{fig:joint_posteriors}c.  
\end{acknowledgements}

%\section{References}\label{references}

%\bibliography{wench_jgr_bib}{}
%\bibliographystyle{agufull08}

>>>>

@@

@@

@@

@@

@@

@@

@@

@@

@@

The magnitudes of $T_{\mathrm{max}}$ and $T_{\mathrm{min}}$ are
dominantly constrained on the low end by our analysis, which is apparent
by the sharp decrease in the frequency of $p(T_{\mathrm{max}}|D)$ below
about $T^\prime_{\mathrm{max}} = 0.5$ (Figure~\ref{fig:joint_posteriors}). 
These results indicate that $T_{\mathrm{max}}$ of at
least $\sim$13.25 MPa km$^{-1}$ is necessary to overcome
topographic stresses resisting reverse and right-lateral slip on the
faults. We find that the highest likely ratio of strike-slip to dip-slip
shear along the Wenchuan earthquake faults is close to 1. This is
similar to the inferences of strain accumulation rates inferred from
squishy block modeling by \citet{loveless2011}, who
concluded that the rate of slip deficit accumulation in the thrust and
dextral senses were approximately equal. It should be noted that the
tectonic stresses we estimate here represent the accumulated stresses
prior to the Wenchuan earthquake, and the relation of these stresses to any
accumulated slip deficit needs to be through a model of strain
accumulation.

@@

@@

\begin{acknowledgements}
All data used in this work are from published sources: \citet{shen2009};
\citet{feng2010}; \citet{qi2011}; \citet{zhang2011}; \citet{fielding2013}. SRTM
data are from CGIAR-SRTM: srtm.csi.cgiar.org. All code used is available at
\url{https://github.com/cossatot/halfspace} and
\url{https://github.com/wenchuan_topo_stress}. We thank the editor, Paul
Tregoning, and two anonymous reviewers for reviews. Funding for
this study was provided by the University of Michigan. Joe Kington wrote the
code to plot Figure \ref{fig:joint_posteriors}c.  
\end{acknowledgements}
